%\VignetteEngine{knitr::knitr}
%\VignetteIndexEntry{flowdurr-datasets}

Vignette for flowdurr datasets---Sample traces from CNRFC, USGS and HSPF
------------------------------------------------------------------------
```{r setup, include=FALSE}
require(knitr)
opts_chunk$set(error=FALSE, tidy=TRUE, dpi=100)
require(flowdurr)
```
>Michael C. Koohafkan   
San Francisco Public Utilities Commission   
Hydrology and Water Systems Modelling   
`r format(Sys.Date(), "%A, %B %d, %Y")`   

Datasets in the flowdurr package
--------------------------------

This vignette documents the four datasets included in package `flowdurr`.

### esptraces
The `esptraces` dataset is a sample set of ESP traces downloaded from <cnrfc.noaa.gov>. The dataset was produced by calling `get_forecast()`. This dataset is updated automatically each package build.
```{r eval=c(2:4)}
esptraces = get_forecast()
data(esptraces)
names(esptraces)
head(esptraces[1:6])
```

### usgstraces
The `usgstraces` dataset is a sample set of USGS traces downloaded by `flowdurr` via the `waterData` package for R. The dataset was produced by calling `get_waterdata()`. This dataset is updated automatically each package build.
```{r eval=c(2:4)}
usgstraces = get_waterdata()
data(usgstraces)
names(usgstraces)
head(usgstraces)
```

### hspftraces
The `hspftraces` dataset is a sample set of HSPF simulation output provided to the package author in CSV format. This dataset was produced by calling `get_hspf("path/to/file.csv")`. This dataset was last updated on 5/19/2014.
```{r eval=c(2:4)}
hspftraces = get_hspf("path/to/file.csv")
data(hspftraces)
names(hspftraces)
head(hspftraces)
```

### workedexample
The `workedexample` dataset contains the variables produced in the `workedexample` demo.
```{r echo=c(2:4), eval=c(1,3:4)}
rm(list=ls())
demo(workedexample)
data(workedexample)
ls()
```
