%\VignetteEngine{knitr::knitr}
%\VignetteIndexEntry{HSPF-example}

Analyzing HSPF simulation output---Another vignette for the flowdurr package for R
----------------------------------------------------------------------------------
>Michael C. Koohafkan   
San Francisco Public Utilities Commission   
Hydrology and Water Systems Modelling   
`r format(Sys.Date(), "%A, %B %d, %Y")`   

```{r setup, include=FALSE}
require(knitr)
opts_chunk$set(error=FALSE, tidy=TRUE, dpi=100)
```

Generating flow duration curves
-------------------------------

We first import the HSPF data. Sample output from a recent HSPF simulation is included in the package. We first load the data, and then reformat and clean up the data. Finally, we produce a spaghetti plot of the traces.
```{r import-data, warning=FALSE, message=FALSE, results='markup'}
require(flowdurr)
# get the data
data(hspftraces)  # could also use get_hspf("path/to/file.csv")
# reformat to water year
hspf.wyear = split_by_wateryear(hspftraces)
head(hspf.wyear[1:10])
# clean the data and plot
hspf.clean = clean_flowdata(hspf.wyear, -1, NA)
hspf.clean = clean_flowdata(hspf.wyear, 0, 1e-5)
plot_all_traces(hspf.clean)
```

We can do the same for the 3-day flow by calculating the 3-day average
timeseries.
```{r avg-flow, warning=FALSE, results='markup'}
hspf.3df = get_flowavg(hspf.clean, 3)
plot_all_traces(hspf.3df)
```

We can also visualize the trace data using boxplots and violin plots.
```{r boxplots, message=FALSE, warning=FALSE, results='markup'}
plot_boxcounts(hspf.clean, flowdurations=c(1,3))
# because these are ggplots, we can add our own aesthetics. Additional aguments
# to geom_boxplot or geom_violin can be passed as well. 
require(ggplot2)
plot_boxcounts(hspf.clean, flowdurations=c(1,3), notch=TRUE) + facet_wrap(~flow.duration)
plot_boxcounts(hspf.clean, flowdurations=1, violin=TRUE, scale='count')
```

We can also plot exceedance counts.
```{r plot-exc, message=FALSE, warning=FALSE, results='markup'}
plot_countexc(hspf.clean, c(150,250,400))
```

Next we compute the peak flow durations.
```{r flow-duration, warning=FALSE, message=FALSE, results='markup'}
hspf.fd = get_peakavg(hspf.clean)
head(hspf.fd)
```

We can visualize the flow durations using histograms or density curves.
```{r plot-fd, warning=FALSE, message=FALSE, results='markup'}
# type ?plot_peakflows to see all plotting options
plot_peakflows(hspf.fd, logscale=TRUE, facet=FALSE)
plot_peakflows(hspf.fd, logscale=TRUE, count=FALSE)
```

We can also calculate the Cunnane plotting positions and plot the empirical cdf.
```{r plotting-positions, message=FALSE, warning=FALSE, results='markup'}
hspf.pp = get_ppositions(hspf.fd)
plot_curves(hspf.fd, hspf.pp, logscale=TRUE)
```

We use the method of L-moments to fit a log Pearson III distribution to the data. For cases where the L-moment estimate is poor, an additional function is provided to refine the parameter estimates using a genetic algorithm.
```{r fittings, warning=FALSE, message=FALSE, results='markup', fig.show='hold', eval=c(1:3,5:9,12)}
# see flowdurr::dlp3 for distribution definition 
hspf.lp3 = get_parameters(hspf.fd, distr='lp3')
## Not run:
hspf.lp3.ref = refine_parameters(hspf.fd, hspf.lp3, distr='lp3')
## End(Not run)
hspf.fit = get_excurves(hspf.lp3, 'lp3')
hspf.fit[seq(from=1, to=nrow(hspf.fit), length.out=10),]
plot_curves(hspf.fd, hspf.pp, hspf.fit, 'HSPF flow durations (LP3)', logscale=TRUE)
## Not run:
hspf.fit.ref = get_excurves(hspf.lp3.ref, 'lp3')
plot_curves(hspf.fd, hspf.pp, hspf.fit.ref, 'HSPF flow durations (LP3)', logscale=TRUE)
## End(Not run)
```

High flow exceedance probabilities
----------------------------------

We return to HSPF traces. Moderately-high flows appear in the 225-375 cfs range:
```{r high-flows, warning=FALSE, results='markup'}
plot_all_traces(hspf.clean) + geom_hline(y=c(225, 375), color='orange')
```

The exceedance probabilities of these flows can be predicted using the 1-day
flow duration curve.
```{r exceedance, warning=FALSE, results='markup'}
# pull high flows from data
flowcols = names(hspf.clean)[names(hspf.clean) != 'GMT' & 
                             names(hspf.clean) != 'WYD']
hspf.highflows = as.matrix(hspf.clean[flowcols])
hspf.highflows = hspf.highflows[!(hspf.highflows < 225 | hspf.highflows > 375)]
#remove NA values
hspf.highflows = hspf.highflows[!is.na(hspf.highflows)]
# restructure to match hspf.fd format
hspf.highflows = as.data.frame(hspf.highflows)
names(hspf.highflows) = '1.day.flow'
hspf.highflows['p.exceedance'] = predict_exceedance(hspf.highflows, hspf.lp3, distr='lp3')[['1.day.flow']]
hspf.highflows
```

Extreme flows ( > 375 cfs) are analyzed similarly.
```{r extreme-flows, warning=FALSE, results='markup'}
hspf.extremeflows = as.matrix(hspf.clean[flowcols])
hspf.extremeflows = hspf.extremeflows[!(hspf.extremeflows < 375)]
hspf.extremeflows = hspf.extremeflows[!is.na(hspf.extremeflows)]
hspf.extremeflows = as.data.frame(hspf.extremeflows)
names(hspf.extremeflows) = '1.day.flow'
hspf.extremeflows['p.exceedance'] = predict_exceedance(hspf.extremeflows, hspf.lp3, distr='lp3')[['1.day.flow']]
hspf.extremeflows
```
