%\VignetteEngine{knitr::knitr}
%\VignetteIndexEntry{plotting-example}

Modifying plot output with ggplot2---yet another vignette for the flowdurr package
----------------------------------------------------------------------------------
>Michael C. Koohafkan   
San Francisco Public Utilities Commission   
Hydrology and Water Systems Modelling   
`r format(Sys.Date(), "%A, %B %d, %Y")` 

```{r setup, include=FALSE}
require(knitr)
opts_chunk$set(error=FALSE, tidy=TRUE, dpi=100, results='markup', warning=FALSE, message=FALSE)
```

The `flowdurr` package provides specialized plots for visualizing flow duration data. The package relies on the popular `ggplot2` package, which provides considerable flexibility for changing plot aesthetics such as colors, labels, and grid lines. Because all plotting functions in `flowdurr` return `ggplot` objects, users can directly modify the plots to optimize readability for presentations, papers or handouts.

We will use the `hspftraces` dataset provided by `flowdurr` to illustrate user modification of plot outputs. We begin by loading and cleaning the data as usual.

```{r import-data}
require(flowdurr)
data(hspftraces)
# reformat to water year
hspf.wyear = split_by_wateryear(hspftraces)
hspf.clean = clean_flowdata(hspf.wyear, -1, NA)
hspf.clean = clean_flowdata(hspf.clean, 0, 1e-5)
hspf.clean = strip_na_cols(hspf.clean)
```

We will work with two plots in this example: a boxplot of the one- and three-day monthly maximum flows, and a spaghetti plot of daily flows.
```{r make-plots, fig.show='hold', echo=c(1:2)}
spaghetti = plot_all_traces(hspf.clean)
box = plot_boxcounts(hspf.clean, flowdurations=c(1,3))
spaghetti
box
```
Let's say we want to use the above plots in a presentation. We will start with the boxplot. In order to improve readability we must
1. Modify the y-axis title
2. increase the text size
3. Change the major and minor grid lines
4. Add annotations
We can do this by simply adding `ggplot2` aesthetics to the plot.

We will start with the spaghetti plot. 
```{r spaghetti-mod1, echo=1:4}
require(ggplot2)
mod.textsize = theme(text=element_text(size=20), axis.text.x=element_text(size=rel(1))) 
mod.yticks = scale_y_continuous('Daily flow (cfs)', minor_breaks=seq(0,600,10), breaks=seq(0,600,100))
spaghettimod = spaghetti + mod.textsize + mod.yticks
spaghettimod
```

We also need to modify the legend to indicate the timespan and total number of traces.
```{r spaghetti-mod2, echo=1:4}
traces = as.numeric(substr(names(hspf.clean)[3:ncol(hspf.clean)], 3, 6))
tlbl = paste(min(traces), '-', max(traces), ' (', length(traces), ' traces)', sep='')
mod.legend = scale_color_manual('', labels=c(tlbl, 'mean', 'median'), 
                                values=c('black', 'red', 'blue'))
spaghettimod = spaghettimod + mod.legend + theme(legend.position='top')
spaghettimod
```

We can easily save high-quality graphics using `ggsave`.
```{r save-plot, eval=FALSE}
ggsave('spaghettiplot.pdf', plot=spaghettimod, width=9, height=6.5)
```

The boxplot requires similar modifications, but requires special treatment to accommodate the log scale on the y-axis.
```{r box-mod1, tidy=FALSE, echo=1:8}
mod.yticks = scale_y_continuous('annual maximum 1-day, 3-day flow (cfs)',
               limits=c(1, 600), breaks = c(1, 5, 10, 50, 100, 500),
               minor_breaks=c(seq(0.01,0.09,0.01), seq(0.1,0.9,0.1), 
			   seq(1,9,1), seq(10,90,10), seq(100,600,100))) 
mod.legend = scale_fill_brewer('', palette='Set1', 
               labels=c("max 1-day flow", "max 3-day flow"))
boxmod = box + mod.yticks + mod.legend + mod.textsize + 
         theme(legend.position='top') + coord_trans(y="log10")
boxmod
```

There is nothing stopping us from fundamentally changing these plots.
```{r restructure-plot}
# check out the structure of a ggplot
names(boxmod)
boxmod$mapping
boxmod$layers
# you can replace layers directly
class(boxmod$layers[[1]])
# change to a jitter plot
boxmod$layers[[1]] = NULL
boxmod + geom_jitter(pch=21)
# but changing aesthetics will reset the legend and color scale
boxmod$mapping[[3]] = NULL
boxmod + geom_jitter(aes(color=flow.duration))
```

It may be easier to just pull the raw data and make a new plot.
```{r remake-plot}
plotdata = boxmod$data
summary(plotdata)
ggplot(plotdata, aes(x=max.flow, fill=flow.duration)) + geom_histogram(position ="dodge") + facet_wrap(~month, scale="free")
```
